---
title: "Speeding up inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Speeding up inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This page is designed to explain how `outerbase` can facilitate fast inference with smart modeling choices.  
```{r setup}
library(outerbase)
```
This is important as the sample size grows, so we will use `500` datapoints.
```{r}
sampsize = 500
d = 8
x = matrix(runif(sampsize*d),ncol=d)
y = obtest_borehole8d(x)
```
The differences get more dramatic with larger sample sizes, however in the spirit of running quickly we will use this sample size.

First setup an `outermod` object.
```{r}
om = new(outermod)
setcovfs(om, rep("mat25pow",8))
knotlist = list();
for(k in 1:d) knotlist[[k]] = seq(0.01,1,by=0.025)
setknot(om, knotlist) #40 knot point for each dim
```

More data should mean more basis functions.
```{r}
p = 250
terms = om$selectterms(p)
```






## Different models

To begin, lets use `?loglik_std` to represent our slow approach.
```{r}
loglik_slow = new(loglik_std, om, terms, y, x) 
logpr_slow = new(logpr_gauss, om, terms)
logpdf_slow = new(lpdfvec, loglik_slow, logpr_slow)
```

`logpdf_slow` can be optimized using `lpdf$optnewton`.
```{r}
logpdf_slow$optnewton()
```
Newton's method involves solving a linear system, thus it takes one step, but is expensive.

`?loglik_gauss` is a `lpdf` model designed for speed.  It is a nice comparison because `loglik_gauss` uses the same model as `loglik_std`, with a few approximations for speed. 
```{r}
loglik_fast = new(loglik_gauss, om, terms, y, x) 
logpr_fast = new(logpr_gauss, om, terms)
logpdf_fast = new(lpdfvec, loglik_fast, logpr_fast)
```


`logpdf_fast` will through an error if you try to use `optnewton`.  This is because it is written so that it never builds a `hess` matrix.
```{r, error=TRUE}
logpdf_fast$optnewton()
```
It is suggested to use `lpdf$optcg` (conjugate gradient) to optimize the coefficients in the fast version.
```{r}
logpdf_fast$optcg(0.001, 100)
```

As an aside, `omp` speed ups are possible, but you need to have correctly compiled with `omp`.  One check is to call the following.  
```{r}
ob = new(outerbase, om, x) 
ob$nthreads
```
If the answer is `1` but you have a multi-threaded processor (most modern processors), things are likely mis-aligned.

You can manually set the number of threads for `lpdf` objects.
```{r}
logpdf_slow$setnthreads(4)
logpdf_fast$setnthreads(4)
```

## Timing

The main cost with models like these is hyperparameter optimization.  Thus will we show the difference in time to build hyperparameters.  Let's save starting points (since they share `om`) for fairness.
```{r}
hyp0_slow = list(para = getpara(logpdf_slow), hyp = gethyp(om))
hyp0_fast = list(para = getpara(logpdf_fast), hyp = gethyp(om))
```

Test points will verify the predictions are equally good with either model, the only difference is speed.
```{r}
xtest = matrix(runif(1000*d),ncol=d) #prediction points
ytest =  obtest_borehole8d(xtest)
```
We will use the unsophisticated ` proc.time` to do some quick timing comparisons.  
```{r}
ptm = proc.time()
opth = BFGS_lpdf(om, logpdf_slow, 
                 rho=hyp0_slow, newt=T)  
t_slow = proc.time() - ptm
pred_slow = new(predictor,loglik_slow)
pred_slow$update(xtest)
yhat_slow = as.vector(pred_slow$mean())
print(t_slow)
```


```{r}
ptm = proc.time()
opth = BFGS_lpdf(om, logpdf_fast, 
                 rho=hyp0_fast, newt=F)  
t_fast = proc.time() - ptm
pred_fast = new(predictor,loglik_fast)
pred_fast$update(xtest)
yhat_fast = as.vector(pred_fast$mean())
print(t_fast)
```
## Comparison of results

And simply plotting the results tells the story: faster inference with no discernible dropoff in quality.  
```{r}
par(mfrow=c(1,2))
rmse_slow = sqrt(mean((ytest-yhat_slow)^2))
hist((ytest-yhat_slow), main=paste("slow method \n rmse:", 
                                    round(rmse_slow,3),
                                   ", time:",
                                   round(t_slow[3],2),'s'),
     xlab = "prediction residuals")
rmse_fast = sqrt(mean((ytest-yhat_fast)^2))
hist((ytest-yhat_fast), main=paste("fast method \n rmse =",
                                      round(rmse_fast,3),
                                   ", time:",
                                   round(t_fast[3],2),'s'), 
     xlab = "prediction residuals")
```


